---
title: "Linking biological data to climate with climatelink"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking biological data to climate with climatelink}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## The problem

You have biological (or clinical, or ecological) data where **every row has its
own place and its own time** — a trapped animal, a blood sample, a nest check, a
hospital admission. You want to know what the climate was like at each of those
places, in the window of time *before* each event, because that is what an
organism actually experiences.

This is a spatiotemporal join, and it's fiddly to do by hand: pull a daily
climate series for each site, line the dates up, and compute rolling summaries
over the right lag window. `climatelink` does it in one call.

## Setup

```{r}
library(climatelink)

# Tell climatelink which Python to use (needs earthengine-api + geemap + pandas),
# and where the Climate_API repo lives (auto-detected if you're inside it).
climate_config(python = "/opt/anaconda3/bin/python3")

# Confirm the backend is healthy.
climate_backend_status()
```

If you have never used Earth Engine on this machine, authenticate once:

```{r}
ee_authenticate(project = "your-gee-project")
```

## A minimal example

```{r}
obs <- data.frame(
  sample_id = c("S001", "S002", "S003"),
  species   = c("Mastomys", "Mastomys", "Rhabdomys"),
  latitude  = c(-26.2678, -26.2678, -33.9249),
  longitude = c( 27.8607,  27.8607,  18.4241),
  date      = as.Date(c("2021-01-15", "2021-06-20", "2021-03-10"))
)

linked <- link_climate(
  obs,
  variables  = c("temperature", "precipitation"),
  windows    = c(7, 30),
  buffer_km  = 10,
  project_id = "your-gee-project"
)

str(linked)
```

Each observation now carries, for example, `tmean_celsius_mean_30d` (mean
temperature over the 30 days before the sample) and `precipitation_mm_sum_30d`
(total rainfall over the same window), plus a `coverage_30d` flag.

## Exposure windows and lags

A *window* of `N` days ends on the observation date and looks back `N` days. If
you want to **exclude** the days immediately around the event (say you believe
the biological response lags by a week), shift the whole window with `lag`:

```{r}
# Summarise days 8–14 before each event (a 7-day window, lagged by 7):
linked_lagged <- link_climate(
  obs, variables = "temperature", windows = 7, lag = 7,
  project_id = "your-gee-project"
)
```

## Choosing aggregations

Defaults are sensible per variable (temperature is averaged and maxed,
precipitation is summed). Override them per raw column:

```{r}
linked_custom <- link_climate(
  obs,
  variables    = "temperature",
  windows      = c(14, 30),
  aggregations = list(tmean_celsius = c("mean", "min", "max"),
                      tmax_celsius  = c("max")),
  project_id   = "your-gee-project"
)
```

## Data quality: always check coverage

`coverage_Nd` is the fraction of each window's days that had data. A value below
`1.0` usually means the point sits on a masked ERA5-Land pixel (common near
coasts and large water bodies). The fix is a spatial buffer, which averages the
surrounding land pixels:

```{r}
# Without a buffer, a coastal point may return NA:
link_climate(obs[3, ], variables = "temperature", windows = 30,
             project_id = "your-gee-project")$coverage_30d   # -> 0

# With a buffer it resolves:
link_climate(obs[3, ], variables = "temperature", windows = 30, buffer_km = 10,
             project_id = "your-gee-project")$coverage_30d   # -> 1
```

## From and to CSV

```{r}
link_climate_csv(
  "field_samples.csv", "field_samples_climate.csv",
  variables = c("temperature", "precipitation"),
  windows   = c(7, 30), buffer_km = 10,
  project_id = "your-gee-project"
)
```

## Next steps

The enriched table is an ordinary data frame — hand it straight to your model:

```{r}
# e.g. does 30-day heat load predict an outcome?
glm(outcome ~ tmean_celsius_mean_30d + precipitation_mm_sum_30d,
    data = linked, family = binomial)
```
